---
title: "Assignment 5 - Statistics II, Hertie School"
output: html_document
---

```{r setup, include = FALSE}
library(foreign)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(summarytools)
library(AER)
library(stargazer)
set.seed(123)
```

<br>

### Task 1 - On the search for plausible instruments [4 points in total]

Come up with a plausible instrument in the following scenarios and justify your choice in 1-3 sentences.

(a)  Scenario 1 [2 pts] <br>
Units: US counties at US House election <br>
Treatment: voter turnout  <br>
Outcome: share of Democrat party <br>

<div style="color:blue">
The researcher can encourage one group to vote by sending reminder text messages, but not to the other. Subjects should be randomly assigned to the encouragement group to ensure that the share of democratic party is not correlated with the instrumental variables. Since the instrument variable will only affect the outcome through treatment, is observable (0 or 1), and is independent from the outcome, it is a plausible instrumental variable. 
</div>

(b) Scenario 2 [2 pts]  <br>
Units: Voters  <br>
Treatment: exposure to Fox News  <br>
Outcome: vote for Republican party <br>

<div style="color:blue">
The position of Fox channel in the cable network affects exposure to Fox News. In a region where Fox News has the default position, viewers will access Fox News more easily. While it is not correlated with the share of democratic party, it can affect the vote for republican party through the mediator, exposure to fox news channel. 
</div>

<br>

### Task 2 - Bert and Ernie's impact on letter recognition [14 points in total]

Suppose we want to estimate the effect of watching an educational television program - Sesame Street - on letter recognition. We might consider implementing a randomized experiment where the participants are preschool children, the treatment of interest is watching Sesame Street, the control condition is not watching, and the outcome is the score on a test of letter recognition. It is not possible here for the experimenter to force children to watch a TV show or to refrain from watching (the experiment took place while Sesame Street was on the air).

```{r, message=F}
dat <- read.dta("sesame.dta")
```

Using the dataset `sesame.dta` where letter recognition is the outcome of interest (`postlet`), regularly watching Sesame Street is the treatment (`regular`) and the encouragement to watch is the instrument (`encour`), do the following:

(a) Discuss the assumptions needed for the identification of the Sesame Street effect and potential violations. [4 pts]

<div style="color:blue">
1. Relevance: encouraging kids to watch the show has a statistically significant effect on whether they watch the show regularly or not. If the encourage does not make them watch it regularly, then the assumption is violated. This can be empirically tested, as the effect of encouragement on treatment is observable and measurable.

2. Ignorability(exogeneity of the instrument): The potential outcome('postlet') and the treatment'regular' are independent of 'encour'. This violation is violated when there are omitted variable bias in the model. That is, if there is a third variable that makes the outcome correlated to Z, the assumption no longer holds. If the encouragement groups are not randomly assigned, the ass of encouragement will address this issue. 

3. Exclusion restriction: the instrument affects outcome only via the treatment. That is, the encouragement will change the outcome only by affecting kid's viewership of the Sesame street. This assumption implies that the ITT(intent-to-treat) effect is 0 for always-takers and never-takers. If the encouragement changes the score of letter recognition directly, not through the treatment, then this assumption is violated.

4. Monotonicity: There are not defiers in the experiment. If kids do the opposite of what they are told, then the assumption is violated.  
</div>

<br>

(b) Illustrate visually the difference in the outcome of interest between the treatment and the control group. [2 pts]

```{r, message=F}
ggplot(dat, aes(x=factor(regular),
                y=postlet,
                fill=factor(encour))) +
  geom_boxplot() + xlab("Regular Sesame") + ylab("Letter Recognition Score") + labs(fill="Encouragement")

```

<br>

(c) What do we know about non-compliance in the data? Give a table or visual representation plus a short description. Also, report a statistic showing whether the encouragement can or cannot be considered a strong instrument. [2 pts]

```{r, message=F, results='asis'}
st_options(footnote=NA)
print(ctable(dat$regular, dat$encour, prop="n"), method="render")
```
<br>
<div style="color:blue">
Assuming that there are no defiers, the cross-table tells us that there are 14 never-takers who were encouraged (E=1) but did not watch regularly (R=0), and 48 always-takers who were not encouraged(E=0) but watched the show regularly (R=1). In total, there are 62 non-compliers in this experiment. 
</div>
<br>

```{r}
summary(lm(regular~encour, dat))
```
<br>
<div style="color:blue">
Linear regression of 'regular' on 'encour' shows us that the relevance of our instrumental variable is statistically significant. The F statistic (50.46), is much greater than 10, which means that the encourgement is a strong instrument. Substantively speaking, 54% of children who were not encouraged watched Sesame, while 90% of children who were encouraged watched Sesame. 
</div>

(d) Provide a naïve estimate of watching on letter recognition, ignoring the encouragement mechanism. Then, use other variables in the data set for a selection-on-observables approach to calculate a less-naïve estimate. [3 pts]

```{r, eval= TRUE}
NAIVE <- lm(postlet~regular, dat)
summary(NAIVE)
```
```{r, eval= TRUE}
NAIVE2 <- lm(postlet~regular+age+sex+prelet+site, dat)
summary(NAIVE2)
```
<br>
<div style="color:blue">
The naive estimates tell us that ... <br>

When not controlling for other variables, NATE1 is 12.69. That is, children who watch Sesame Street regularly scores 12.69 points higher that children who did not. <br>
<br>
Controlling for age, sex, pre-score of letter recognition (to control for baseline bias) and site (to control for the difference in effect depending on where children watch it), NATE is 10.45. That is, children who watch Sesame regularly scores 10.45 points higher than the children who did not.<br>
<br>
This estimate is naive because it does not take the effect of an unobserved variable into consideration. That is why we need to use instrumental variables!
</div>

(e) Provide estimates for the ITT and the LATE and give a brief substantive interpretation of both results! [3 pts]
  
```{r, eval= TRUE}
ITT <- lm(postlet~encour, dat)
summary(ITT)
```
<br>
<div style="color:blue">
Intent-to-treat (ITT) effect is the effect of the instrument regardless of the actual treatment. Under the assumptions of monotonicity and exclusion restriction, the ITT can be interpreted as local average treatment effect (LATE) for compliers, which is the proportion of compliers in the sample multiplied by ITT of compliers. Here, the effect of encouragement on the outcome (ITT) is 2.876, howevever it is not statistically significant. 
<br>

LATE can be calculated using the Wald Estimator, which is the covariance between Y and Z divided by covariance between D and Z, where <br>
cov(Y,Z) = E(Y|Z=1) - E(Y|Z=0) #as given by the regression of the outcome on the instrument (task 2e) <br>
cov(D,Z) = E(D|Z=1) - E(D|Z=0) #as calculated using the cross-table for compliers (task 2c) <br>
<br>
cov(Y,Z) = 27.796  - 24.920 = 2.876 <br>
cov(D,Z) = (138/152) - (48/88) = 0.3624402<br>
<br>
LATE = 2.876/0.3624402 = 7.935<br>
<br>
This can be calculated using ivreg function as well.
```{r}
summary(ivreg(postlet~regular | encour, data=dat))
```

</div>

<br>

### Task 3 - Simulating a flawed instrument [3 points in total]

Go back to the simulation code from the Lab and refine it such that Z is a flawed instrument as it is also connected with Y through a variable other than X. Then, show how these modifications change the estimates of the original setup!  [3 pts]

```{r}
# generate xStar and c and using a common variance
xStarAndC <- MASS::mvrnorm(1000, c(20, 15), matrix(c(1, 0.5, 0.5, 1), 2, 2))
xStar <- xStarAndC[, 1]
c <- xStarAndC[, 2] 

# For z to be an invalid instrument, C has to be related to both x and z. 
z <- rnorm(1000) + 4*c
x <- xStar + z + 3*c

# z is no longer a valid instrument because outcome y is affected by v, which is correlated with z. 
y <- 1 + x + c + rnorm(1000, 0, 0.5) 
```
```{r}
summary(lm(x~z)) # F statistic is much lower, meaning that our instrument is no longer valid. 
```
<br>




