---
title: "230420_Replication Paper Code"
author: "Sophie Giulini"
date: "4/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading libraries

```{r warning=FALSE, message=FALSE}

set.seed(42) #for consistent results

library(dplyr) # to wrangle our data
library(tidyr) # to wrangle our data - gather()
library(ggplot2) # to render our graphs
library(haven) # to load our .dta files for the assignment
library(plm) # to run our fixed effects models
library(kableExtra) # to render better formatted tables
library(stargazer) # for formatting your model output
library(tidyverse)
library(lmtest)# for calculating clustered standard errors 
library(MatchIt)# for matching

```

Loading data

```{r warning=FALSE, message=FALSE}
vote_centers_data <- read_dta("Rottinghaus_Cortina_Vote Centers and Turnout by Election Type in Texas_Data.dta")
vote_centers_data

#creating sub sets with variables, which are known and required for the data replication
Selected_data <-vote_centers_data%>%select(year, turnout, election, county,medianinc_rec, bachelor_higher, presidential, constitutional, midterm, totalpop_rec, votecenter)%>%arrange(county,election)
Selected_data

length(unique(Selected_data$county))
#Sample size of 254 

parllel_trend_assumption_data<-read_dta("Rottinghaus_Cortina_Vote Centers and Turnout by Election Type in Texas_Parallel Assumption Data.dta")
parllel_trend_assumption_data

names(vote_centers_data) # to check the names of the variables in our data
```

#Table 1 

To replicate table 1, a Difference-in-Difference (DD) fixed effects model is built with help of the plm function. 
To analyse the DD effect of votecenters on turnout for the different election cycles, one can create dummy variables, where 1 represents the voting year after, and 0 the voting year before the implementation of votecenters.

```{r warning=FALSE, message=FALSE}

#Creating dummy variables for the different election years

Selected_data %>% mutate(pt=case_when(
year == 2016 ~ "1",
year==2012~ "0"
  ))

Selected_data<-Selected_data %>% mutate(pt)

vote_centers_data %>% mutate(pt=case_when(
year == 2016 ~ "1",
year==2012~ "0"
  ))

vote_centers_data %>% mutate(mt=case_when(
  year== 2018 ~ "1",
  year== 2010 ~ "0"
  ))

vote_centers_data %>% mutate(ct=case_when(
  year == 2017 ~ "1",
  year == 2009 ~ "0"
  ))

vote_centers_data<-vote_centers_data %>% mutate(pt, mt, ct)

names(vote_centers_data)

#Modeling a fixed effects function with DD with help of plm 

reg1 <- plm(turnout ~ votecenter + pt + votecenter * pt + medianinc_rec + bachelor_higher + totalpop_rec , data=vote_centers_data, index = c("county"), model = "within")
reg1

reg1 <- plm(turnout ~ votecenter + pt + votecenter * pt + medianinc_rec + bachelor_higher+totalpop_rec, data=Selected_data, index = c("county"), model = "within")
reg1

reg2 <- plm(turnout ~ votecenter + mt + votecenter * mt + medianinc_rec + bachelor_higher + totalpop_rec, data=vote_centers_data, index = c("county"), model = "within")
reg2

reg3 <- plm(turnout ~ votecenter + ct + votecenter * ct + medianinc_rec + bachelor_higher + totalpop_rec, data=vote_centers_data, index = c("county"), model = "within")
reg3

stargazer(reg1,reg2,reg3,type= "html", out="table1.txt",
          omit.stat=c("LL","ser","f"), no.space=TRUE)

#Calculate clustered standard errors with help of coeftest

coeftest(reg1,vcov. = NULL, df=NULL)
coeftest(reg2,vcov. = NULL, df=NULL)
coeftest(reg3,vcov. = NULL, df=NULL)
```

#Table 2

```{r warning=FALSE, message=FALSE}
# Creating dummy variables for the different time cycles 

ct1 <- case_when(
  vote_centers_data$year %% 2011 == 1 ~ "1",
  vote_centers_data$year %% 2009 == 0 ~ "0"
)

vote_centers_data %>% mutate(ct1=case_when(
  vote_centers_data$year %% 2011 == 1 ~ "1",
  vote_centers_data$year %% 2009 == 0 ~ "0"
  ))


ct2<-case_when(
  vote_centers_data$year %% 2017 == 1 ~ "1",
  vote_centers_data$year %% 2011 == 0 ~ "0"
  )
vote_centers_data %>% mutate(ct2=case_when(
  vote_centers_data$year %% 2017 == 1 ~ "1",
  vote_centers_data$year %% 2011 == 0 ~ "0"
  ))

ct3<-case_when(
  vote_centers_data$year %% 2013 == 1 ~ "1",
  vote_centers_data$year %% 2015 == 0 ~ "0"
  )

vote_centers_data %>% mutate(ct3=case_when(
  vote_centers_data$year %% 2013 == 1 ~ "1",
  vote_centers_data$year %% 2015 == 0 ~ "0"
  ))

#Modeling unit-fixed effects function with DD, with help of plm

reg4 <- plm(turnout ~ votecenter + ct1 + votecenter * ct1 + medianinc_rec + bachelor_higher + totalpop_rec, data=vote_centers_data, index = c("county"), model = "within")
reg4

reg5 <- plm(turnout ~ votecenter + ct2 + votecenter * ct2 + medianinc_rec + bachelor_higher + totalpop_rec, data=vote_centers_data, index = c("county"), model = "within")
reg5

reg6 <- plm(turnout ~ votecenter + ct3 + votecenter * ct3 + medianinc_rec + bachelor_higher + totalpop_rec, data=vote_centers_data, index = c("county"), model = "within")
reg6

# Calculating clustered stard errors 

coeftest(reg4, vcov. = NULL, df=NULL)
coeftest(reg5, vcov. = NULL, df=NULL)
coeftest(reg6, vcov. = NULL, df=NULL)

stargazer(reg4, reg5, reg6, type= "html", out="table1.txt",
          omit.stat=c("LL","ser","f"), no.space=TRUE)

```

#Table 3 

To replicate table 3, doing a Dif-in-Dif analysis on matched data, we build three data subsets for the different election cyclys, which are Presidential, Constitutional and Midterm elections. 
Within the subsets, we match the data. 

```{r warning=FALSE, message=FALSE}
#Creating data sub sets for the different election cycles
presidential_d <-Selected_data%>%filter(election=="presidential"&year==2016|year==2012)%>%mutate(time_p(ifelse(year==2016,1,0))
                                                                                            
midterm_d <- Selected_data%>%filter(election=="midterm"&year==2018|year==2010)%>%mutate(time_m(ifelse(year==2018,1,0))   
                                                                                        
constitutional_d <- Selected_data%>%filter(election=="constitutional"&year==2017|year==2009)%>%mutate(time_c(ifelse(year==2017,1,0))
```                                                                                                     
                                                                                                      
Propensity score matching on co-variates for the three subsets with matchit 

```{r warning=FALSE, message=FALSE}

#Estimate logit model
logitmodel <- glm(votecenter ~ medianinc_rec + bachelor_higher + totalpop_rec,
            family = binomial(), 
            data = presidential_d)

stargazer(logitmodel, type = "html")

```

```{r warning=FALSE, message=FALSE}
#Predicted probabilities
prs_df <- data.frame(pr_score = predict(logitmodel, type = "response"),
                     votecenter = logitmodel$model$votecenter)

#Plot probablity
prs_df %>%
  ggplot(aes(x = turnout, fill = factor(votecenter))) +
  geom_density(alpha = 0.5) +
  labs(x = "Propensity Score Distribution: Treatment and Control Groups",
       fill = "Establishment of votecenters")
````

```{r warning=FALSE, message=FALSE}
#Non-exact matching: 
matched_presidential <- matchit(votecenter ~  medianinc_rec + bachelor_higher + totalpop_rec, data = presidential_d,  
   method = "nearest", 
   ratio = 1, 
   replace = TRUE,
   data = XXX))

summary(matched_presidential)
```



                                                                                                      
```

# Extension


